What defines big team science?

### 2025-06-17 manuscript_definition.Rmd
updating for publication

  Funding: No funding was obtained for this project. #we staying with this even though it helped with coding errors?

  Artificial intelligence: No AI resources were used for this manuscript.

@@

Scientific discovery has increasingly become a collaborative process,
with the scale and scope of team science have dramatically expanded in
recent years [@council2015]. Collaboration in scientific endeavors
involves multiple researchers (potentiality) at multiple institutions to
communicate and work together to advance knowledge in their chosen
field(s). This unique composition of a collaboration for each project is 
dependent on the skill sets, hypotheses, and perspectives of collaborators 
involved. A key strength of collaboration lies in its flexibility, allowing it 
to adapt to the needs of the project and the researchers themselves.

While collaboration is not new in science, interest in "team science" is
growing as individual researchers seek an interdisciplinary approach to
research or bring on more students to their project. Team science is
often defined as groups of researchers with various expertise working
together to investigate complex problems [@fiore2008]. Unlike general
collaboration, team science involves structured roles, coordinated
workflows, and shared resources to address challenges that would be
difficult for individual researchers or one small team to solve
independently.

@@

Beyond replication concerns, the credibility movement has mirrored calls
for diversification or de-WEIRDing (e.g., Western, Educated,
Industrialized, Rich, and Democratic) scientific research [@henrich2010;
@newson2021; @rad2018] by improving representation in research samples.
Like the large-scale studies in Physics [@aphilos2021; @castelnovo2018]
and Biology [@collins2003], the Social Sciences struggle to represent
the breadth of humanity across both researcher and population
characteristics. Now, grassroots organizations, such as the
Psychological Science Accelerator [@moshontz2018], ManyBabies
(<https://manybabies.github.io/>), NutNet (<https://nutnet.org/>),
DRAGNet (<https://dragnetglobal.weebly.com/>), and IceCube
(<https://icecube.wisc.edu/>) can begin to tackle these issues by
recruiting research labs from all over the globe to provide diversity in
geographic, linguistic, and researcher representation. Publications have
examined the global understanding of morality, face processing, COVID-19
information signaling, and more [@bago2022; @dorison2022; @jones2021;
@psychologicalscienceacceleratorself-determinationtheorycollaboration2022;
@vanbavel2022; @wang2021]. While these organizations and one-time groups
for BTS studies have provided an incredible wealth of data for the
scientific community, we do not yet know exactly how to define BTS -
which is generally termed "an unusually large number of collaborators"
[@forscher2022a; @coles2022].

@@

-   Research Question 1: Exploring historical and current publication
    values, what should define big team science versus team science?
    -   Question 1A: What number of authors and institutional
        affiliations should designate the differences between team
        science and big team science?
    -   Question 1B: Using the definition from 1A, are there changes in
        the number of publications over time?
-   Research Question 2: How has the diversity of those involved in team
    science and big team science changed over time?

@@

```{r total-papers}
total_papers <- 97532104 # the entire database
total_papers_past1970 <- 62966549 # past 1970, in our article types
distinct_authors_past1970 <- 53622443 # all authors

@@

The total number of papers included in the Scopus database at the time
of this analysis was `r total_papers`. `r total_papers_past1970`
articles were included past 1970 in the defined article types, which
included `r distinct_authors_past1970` distinct authors. We then
filtered the data to include only teams, which was defined as two
authors from at least two institutions. The total number of papers for
team projects was `r team_author_aff_papers` and
`r team_author_aff_author` distinct authors. The data was then
classified into subject areas by paper, which lead to missing data. The
final number of papers included was `r team_author_aff_papers_noNA` with
`r team_author_aff_author_noNA` distinct authors. The dataset was
curated to include one row per author, paper, and subject area (i.e.,
long format [@wickham2007]) which included
`r team_author_aff_subject_noNA` total rows of data.

@@

@@

Supplemental Table \@ref(tab:big-teams-table) includes the number of
distinct authors and papers for each subject area by overall teams and
big teams using our 90th percentile definition. The total number of
distinct authors for big team papers was `r big_teams_papers` with
`r big_teams_authors` distinct authors. In RQ2, we split the big team
data into small (11-49), medium (50-99), and large big team (100+)
groupings for convenience to display/analyze geopolitical regions. The
table shows the number of authors and papers for those analyses.

@@

For analyzing changes across time, we split the data into team science
projects (2-10 authors, 2-5 affiliations) and BTS projects (as defined
above, 11+ authors, 6+ affiliations). The number of papers found in
Scopus across time for each subject area are displayed in Figure
\@ref(fig:fig-paper-time). The visual results indicated that the number
of team science papers was increasing the most in physical sciences for
all manuscripts, followed by life and health sciences, and the last is
social sciences. Examining only BTS projects shows that the rate is also
increasing across time. All teams appear to start increasing in the
1990s, while BTS projects do not start increasing off floor effects
until past 2000. The health and life sciences show the largest increases
across time in big teams with the smallest trend in the social sciences.

```{r fig-paper-time, include = TRUE, fig.cap = "The number of manuscripts across time for all team science papers (left) and big team science papers (right).", fig.width = 10, fig.height = 6}
temp <- DF$author_paper_year %>% 
  filter(SubjectArea != "Multidisciplinary") %>% 
  filter(Year < 2025) %>% 
  mutate(SubjectArea = gsub("Sciences", "", SubjectArea))

p1 <- ggplot(temp %>% 
         filter(team_size == "regular"), 
         aes(x = Year, y = Number_Papers, 
             colour = SubjectArea)) + 
  geom_point() + 
  theme_minimal() + 
  scale_color_discrete(name = "Subject Area") + 
  theme(legend.position = "bottom",
        axis.text.y = element_text(angle = 45, hjust = 1)) + # Angles the y-axis text
  ylab("Number of Papers") + 
  ggtitle("Team Science Projects") +
  theme(legend.title = element_text(size = 15),
    legend.text = element_text(size = 12))

@@

Using the `minpack.lm` library [@elzhov2023], we calculated the
exponential rate of growth for team science and BTS projects, and these
results are shown in Supplemental Figure \@ref(fig:rate-estimates-fig). All growth rate
confidence intervals excluded zero, indicating an exponential increase
in the number of team papers over time. BTS growth rates were always
higher than their team science counterparts, but the 95% confidence
intervals for the growth estimate overlapped for all statistics.
Therefore, the growth trends, while visually appearing to be different,
were likely similar for each subject area and team size when examined by
estimating exponential growth statistics.

@@

corr.DF <- bind_rows(corr.results) %>% 
  mutate(group = rep(group, each = 8)) %>% 
  mutate(SubjectArea = c(
    rep(subjarea, each = 2),
    rep(subjarea, each = 2)
    )) %>% 
  filter(term != "a") %>% 
  mutate(conf.low = estimate-2*std.error,
         conf.high = estimate+2*std.error) %>% 
  mutate(group = factor(group, 
                        levels = c("regular", "big"),
                        labels = c("Team Science", "BTS")))

@@

```{r fig-career, include = TRUE,  warning = FALSE, message = FALSE, fig.cap = "Average career length for big-team science authors. Larger dots indicate more variability in career length for authors by averaging the standard deviation in career length for each manuscript within a year. The data has been filtered to at least 10 publications in a year for this graph.", fig.width=10, fig.height=6}
ggplot(DF$career_time %>% 
         filter(SubjectArea != "Multidisciplinary") %>% 
         filter(num_papers >= 10) %>% 
         filter(Year < 2025) %>% 
         mutate(Team = factor(Team, 
                              levels = c("Regular", "Big"),
                              labels = c("Team Science", "BTS"))), 
       aes(Year, avg_career, color = Team)) +
  theme_classic(base_size = 15) + 
  geom_point(aes(size = sd_career)) +
  facet_wrap(~SubjectArea) + 
  theme(legend.position = "bottom") + 
  ylab("Average Career Length") +
  xlab("Year of Publication") + 
  scale_color_manual(values = c("#D55E00", "#0072B2"), name = "Team Size") + # Red and Blue from RdBu ends
  guides(size = "none")
  #scale_size_continuous(name = "Number Papers")

@@

Figure \@ref(fig:fig-career) portrays the average career length for
authors involved in team science and BTS projects over time. Career
length was defined as the year of first publication minus the current
year, and higher numbers mean longer careers. The general pattern for
team science and BTS projects is a decrease in average career length
over time. However, it appears that, in at least the last two decades,
BTS projects average a longer career length than team science projects.
This trend is visually consistent across all four subject areas
examined.

@@

All slopes for both the average ($b_M$) and standard deviation
($b_{SD}$) of publication count were significantly different from zero,
indicating meaningful change over time in the types of researchers
contributing to publications across subject areas and team sizes (see
Figure \@ref(fig:fig-heatmap)). Most subject areas differed
significantly from one another within their team size, with the
exception of Health Sciences and Physical Sciences for team science
teams, whose $b_M$ values overlapped in their confidence intervals.
Across the remaining comparisons, Life Sciences showed the steepest
decline in average publication count over time for regular teams ($b_M$
= -0.0588), suggesting a shift toward including authors with fewer
publications. In contrast, the smallest change in publication count was
observed in Physical Sciences for big teams ($b_M$ = -0.0005),
indicating some stability scholar publication count when examining
diversity. Standard deviation slopes ($b_{SD}$) were generally low in
magnitude, with both positive and negative values depending on subject
area. This suggests some variation in the diversity of publication rates
across disciplines, with no uniform pattern of increasing or decreasing
diversity.

All subject areas showed significant differences between BTS and team
science teams in both $b_M$ and $b_{SD}$, as indicated by
non-overlapping confidence intervals. Team science teams consistently
exhibited stronger negative slopes for average publication count than
BTS teams, reflecting a more pronounced trend toward authors with fewer
publications appearing over time. This suggests that smaller teams are
increasingly composed of researchers with lower overall publication
counts, whereas BTS teams show a more muted shift. Effect sizes for
these models were smaller than those observed for career length, with
$R^2$ values ranging from .0335 to .0977. The strongest association was
observed in the Physical Sciences for big teams ($R^2$ = .0977), though
all models showed low-to-moderate predictive ability.

@@

Geographic patterns in authorship differed notably between BTS and team
science publications, as shown in the publication maps and mosaic charts
(Supplemental Figure \@ref(fig:fig-map-both)). BTS publications were overwhelmingly
concentrated in high-income countries, particularly the United States,
Western European nations (e.g., Germany, the United Kingdom, France, and
the Netherlands), and East Asian countries (e.g., China, Japan, and
South Korea). In contrast, team science publications showed broader
geographic distribution, with relatively higher representation from
Latin America (e.g., Brazil, Mexico), South and Southeast Asia (e.g.,
India, Pakistan, Indonesia), and parts of Africa and the Middle East.
While both team types were led by traditionally defined Global North
institutions, the mosaic charts revealed that team science teams
included a more diverse range of countries contributing at moderate
levels. These patterns suggest that although BTS involves international
collaborations, it remains more centralized in historically dominant
research regions, whereas team science may offer relatively greater
global inclusivity at a smaller scale.

To understand the change in representation diversity, we examined if the
number of regions in a publication is predicted by the year of
publication. Increasing diversity would be represented by a positive
slope, while decreasing diversity would be represented by a negative
slope. All slopes predicting geopolitical diversity over time were
significantly different from zero, indicating small but non-zero changes
in the number of regions represented on publications across disciplines
and team types. Additionally, all slopes differed significantly between
BTS and team science publications, suggesting distinct patterns in the
evolution of international collaboration. Within BTS publications, Life
Sciences and Social Sciences showed statistically indistinguishable
trends in regional diversity over time, as did Social Sciences and
Health Sciences. In contrast, all other within-BTS comparisons differed
significantly. For team science publications, all four disciplines
showed significantly different slopes, although the magnitudes of these
differences were relatively small. Overall, the results suggest modest
increases in geopolitical diversity in most disciplines, with a small
decline observed only in Physical Sciences within BTS publications ($b$
= -0.0129). Despite small effect sizes (all $R^2 < .012$), the
consistent differences between BTS and team science teams point to
structural differences in how global participation is evolving across
large-scale versus more traditional collaborations.

@@

Across all decades and team sizes, North America and Northwestern Europe
consistently made up the majority of corresponding authors, as shown in
Figure \@ref(fig:fig-author-gpe). This pattern held even as total
team size increased, though the proportion of corresponding authors from
other regions (e.g., Asia, Latin America and the Caribbean, and Africa)
showed gradual increases over time. For team science teams (2–10
authors), the dominance of North America and Western Europe in
corresponding author roles was particularly pronounced. In contrast, for
very large teams (100+ authors), regional diversity appeared somewhat
more balanced, with more visible contributions from Asia and other
regions among both corresponding and non-corresponding authors. However,
visual inspection suggests that corresponding author sets remained less
regionally diverse than the rest of the author team, particularly in
earlier decades. While representation from regions like Africa and Latin
America grew slightly among non-corresponding authors, they remained
minimally represented in lead authorship positions. Notably, Eastern
Asia’s representation increased more substantially over time, especially
in teams with 50 or more authors. These visual trends suggest persistent
regional disparities in leadership roles within scientific publications,
despite increasing global collaboration. Quantitative effect sizes
(Cramer’s *V*) and confidence intervals are reported in the following
section to determine the importance of these observed differences.

@@

This study expands on prior efforts to characterize Big Team Science
(BTS) by providing a systematic, field-wide analysis of authorship
composition across time, team size, and geography. While BTS efforts
have been increasingly promoted as vehicles for collaboration, scale,
and rigor [@adams2012; @uhlmann2019], questions remain about who gets
included, who leads, and how equitably credit is distributed. A key
contribution of our study is that it is the first to propose a
data-driven operationalization of "big" teams: defined here as
publications with 11 or more authors and contributions from at least 6
institutions, grounded in the empirical distribution of team sizes and
affiliations across millions of papers. By comparing BTS publications
with Team Science publications across four major scientific domains, we
clarify how BTS compares with traditional team science and how it is
evolving. All areas of research show growth in the number of
publications and authors included on manuscripts, replicating previous
investigations [@sinatra2015; @wuchty2007; @hunter2008].

Next, we find that early-career researchers are increasingly represented
in both BTS and smaller teams. Across all disciplines, the average
career length of authors decreased significantly over time. Team Science
teams exhibited even steeper declines in both average and variability of
career length than big teams, suggesting that smaller teams may be an
especially important entry point for early-career researchers. These
trends echo broader shifts in academia’s incentive structures, where
publishing early and often is increasingly required for career
advancement [@milojevic2014; @larivière2015]. Publication counts showed
similar but smaller effects. The average number of publications per
author declined over time in both big and small teams, with smaller
teams again showing more pronounced shifts. These findings support
claims that collaborative science is no longer dominated exclusively by
elite or high-output researchers [@milojevic2014], but may instead be
expanding to include contributors with more varied publication
histories.

Our findings build on previous research by also examining diversity in
author seniority and geopolitical affiliation. The growing participation
of early-career scholars over time suggests that big team science may be
increasingly accessible to a broader range of researchers, not just
senior or established scientists. This trend is interesting given the
challenges BTS projects can pose for non-permanent researchers: slow
publication timelines, uncertain publication outcomes, and fewer
incentives for non-corresponding authors. Yet, large teams may allow for
more distributed workloads and reduced individual time investment, which
could make them appealing even for early-career researchers. Moreover,
prior work has shown that publications from larger teams tend to receive
more citations and have broader impact [@larivière2015], which may
further incentivize early-career involvement despite the structural
risks. Globalization, the internet, and the focus on interdisciplinary
research are potentially driving forces behind our results, but,
hopefully, the results also point to a decline in scientific gatekeeping
[@lu2007; @siler2015].

Our results confirm and extend prior observations that both Team Science
and BTS are disproportionately concentrated in high-resource, highly
networked regions, namely, North America and Western Europe [@singh2023;
@adams2012; @sugimoto2017]. However, this study offers a more nuanced
picture. We observed modest increases in the geographic diversity of
authorship. Yet, lead authorship remained concentrated in a relatively
narrow set of regions. These findings parallel previous critiques of
global equity in scientific collaboration, where authors from the Global
South are often included in co-authorship lists but remain
underrepresented in leadership roles [@chan2011; @sumathipala2004].
Though Cramer’s *V* values reflecting geographic imbalance decreased
over time, especially for small and mid-sized teams, some asymmetries
persist in large teams, reinforcing concerns about exclusion even within
globally scoped research efforts [@abimbola2019].

@@

The limitations for this research are tied to the curation of the Scopus
dataset: the correct author affiliations, the correct author publication
information, and the correctly marked geopolitical entity. Scopus is a
carefully curated and large dataset, but these limitations must be kept
in mind when interpreting the results. Publication language diversity
was not investigated, and a previous study indicates that most
publications in big databases are in English [@albarillo2014].
Certainly, publications in non-English languages would improve the
statistics on diversity in scientific publishing - but the English
language barrier likely exists regardless of inclusion in databases
[@meneghini2007; @ramírez-castañeda2020].

>>>>

  Funding: No funding was obtained for this project.

  Artificial Intelligence: Artificial Intelligence tools (e.g., ChatGPT) were used to assist in debugging and resolving coding errors.

@@

Scientific discovery has increasingly become a collaborative process,
with the scale and scope of team science have dramatically expanded in
recent years [@council2015]. Collaboration in scientific endeavors
involves multiple researchers (potentiality) at multiple institutions to
communicate and work together to advance knowledge in their chosen
field(s). This unique composition of a collaboration for each project is 
dependent on the skill sets, hypotheses, and perspectives of collaborators 
involved. A key strength of collaboration lies in its flexibility, allowing it 
to adapt to the needs of the project and the researchers themselves.  While collaboration is not new in science, interest in "team science" is
growing as individual researchers seek an interdisciplinary approach to
research or bring on more students to their project. Team science is
often defined as groups of researchers with various expertise working
together to investigate complex problems [@fiore2008]. Unlike general
collaboration, team science involves structured roles, coordinated
workflows, and shared resources to address challenges that would be
difficult for individual researchers or one small team to solve
independently.

@@

Beyond replication concerns, the credibility movement has mirrored calls
for diversification or de-WEIRDing (e.g., Western, Educated,
Industrialized, Rich, and Democratic) scientific research [@henrich2010;
@newson2021; @rad2018] by improving representation in research samples.
Like the large-scale studies in Physics [@aphilos2021; @castelnovo2018]
and Biology [@collins2003], the Social Sciences struggle to represent
the breadth of humanity across both researcher and population
characteristics. Now, grassroots organizations, such as the
Psychological Science Accelerator (<https://psysciacc.org>), ManyBabies
(<https://manybabies.github.io/>), NutNet (<https://nutnet.org/>),
DRAGNet (<https://dragnetglobal.weebly.com/>), and IceCube
(<https://icecube.wisc.edu/>) can begin to tackle these issues by
recruiting research labs from all over the globe to provide diversity in
geographic, linguistic, and researcher representation. Publications have
examined the global understanding of morality, face processing, COVID-19
information signaling, and more [@bago2022; @dorison2022; @jones2021;
@psychologicalscienceacceleratorself-determinationtheorycollaboration2022;
@vanbavel2022; @wang2021]. While these organizations and one-time groups
for BTS studies have provided an incredible wealth of data for the
scientific community, we do not yet know exactly how to define BTS: which is generally termed "an unusually large number of collaborators"
[@forscher2022a; @coles2022].

@@

-   Research Question 1: Exploring historical and current publication
    values, what should define big team science versus traditional team science?
    -   Question 1A: What number of authors and institutional
        affiliations should designate the differences between traditional team science and big team science?
    -   Question 1B: Using the definition from 1A, are there changes in
        the number of publications over time?
-   Research Question 2: How has the diversity of those involved in traditional team
    science and big team science changed over time?

@@

```{r total-papers}
num <- function(x){
  format(x, big.mark = ",", scientific = FALSE)
}

total_papers <- 97532104 # the entire database
total_papers_past1970 <- 62966549 # past 1970, in our article types
distinct_authors_past1970 <- 53622443 # all authors

@@

The total number of papers included in the Scopus database at the time
of this analysis was `r num(total_papers)`. `r num(total_papers_past1970)`
articles were included past 1970 in the defined article types, which
included `r num(distinct_authors_past1970)` distinct authors. We then
filtered the data to include only teams, which was defined as two
authors from at least two institutions. The total number of papers for
team projects was `r num(team_author_aff_papers)` and
`r num(team_author_aff_author)` distinct authors. The data was then
classified into subject areas by paper, which lead to missing data. The
final number of papers included was `r num(team_author_aff_papers_noNA)` with
`r num(team_author_aff_author_noNA)` distinct authors. The dataset was
curated to include one row per author, paper, and subject area (i.e.,
long format [@wickham2007]) which included
`r num(team_author_aff_subject_noNA)` total rows of data.

@@

We applied a consistent definition of Big Team Science (BTS) across all four research domains in our analysis. As shown in Figure \@ref(fig:fig-stats-authors), the 90th percentile of team size is consistent across subject areas, varying by only four authors (e.g., from 7 to 11), which supports the use of a unified threshold. Over time (Figure \@ref(fig:fig-paper-time)), we observe increasing publication counts in all fields. While social sciences currently lag behind in both volume and growth rate, we anticipate continued growth as more infrastructure and funding are directed toward collaborative efforts in this area. Unlike fields such as health sciences, which often benefit from greater financial resources and institutional support (e.g., through IRBs and clinical networks), social science collaborations face unique structural barriers that may slow their expansion. Additionally, defining BTS consistently across fields is methodologically important because many publications are interdisciplinary and assigned to multiple subject categories. A unified operationalization allows for clearer comparisons and avoids arbitrary distinctions between overlapping research areas.

@@

Supplemental Table \@ref(tab:big-teams-table) includes the number of
distinct authors and papers for each subject area by overall teams and
big teams using our 90th percentile definition. The total number of
distinct authors for big team papers was `r num(big_teams_papers)` with
`r num(big_teams_authors)` distinct authors. In RQ2, we split the big team
data into bins using 11-49 authors, 50-99 authors, and 100+ authors groupings for convenience to display/analyze geopolitical regions. The
table shows the number of authors and papers for those analyses.

@@

For analyzing changes across time, we split the data into traditional team science
projects (2-10 authors, 2-5 affiliations) and BTS projects (as defined
above, 11+ authors, 6+ affiliations). The number of papers found in
Scopus across time for each subject area are displayed in Figure
\@ref(fig:fig-paper-time). The visual results indicated that the number
of traditional team science papers was increasing the most in physical sciences for
all manuscripts, followed by life and health sciences, and the last is
social sciences. Examining only BTS projects shows that the rate is also
increasing across time. All teams appear to start increasing in the
1990s, while BTS projects do not start increasing off floor effects
until past 2000. The health and life sciences show the largest increases
across time in big teams with the smallest trend in the social sciences.

```{r fig-paper-time, include = TRUE, fig.cap = "The number of manuscripts across time for all traditional team science papers (left) and big team science papers (right).", fig.width = 10, fig.height = 6}
temp <- DF$author_paper_year %>% 
  filter(SubjectArea != "Multidisciplinary") %>% 
  filter(Year < 2025) %>% 
  mutate(SubjectArea = gsub("Sciences", "", SubjectArea))

p1 <- ggplot(temp %>% 
         filter(team_size == "regular"), 
         aes(x = Year, y = Number_Papers, 
             colour = SubjectArea)) + 
  geom_point() + 
  theme_minimal() + 
  scale_color_discrete(name = "Subject Area") + 
  theme(legend.position = "bottom",
        axis.text.y = element_text(angle = 45, hjust = 1)) + # Angles the y-axis text
  ylab("Number of Papers") + 
  ggtitle("Traditional Team Science Projects") +
  theme(legend.title = element_text(size = 15),
    legend.text = element_text(size = 12))

@@

Using the `minpack.lm` library [@elzhov2023], we calculated the
exponential rate of growth for traditional team science and BTS projects, and these
results are shown in Supplemental Figure \@ref(fig:rate-estimates-fig). All growth rate
confidence intervals excluded zero, indicating an exponential increase
in the number of team papers over time. BTS growth rates were always
higher than their traditional team science counterparts, but the 95% confidence
intervals for the growth estimate overlapped for all statistics.
Therefore, the growth trends, while visually appearing to be different,
were likely similar for each subject area and team size when examined by
estimating exponential growth statistics.

@@

corr.DF <- bind_rows(corr.results) %>% 
  mutate(group = rep(group, each = 8)) %>% 
  mutate(SubjectArea = c(
    rep(subjarea, each = 2),
    rep(subjarea, each = 2)
    )) %>% 
  filter(term != "a") %>% 
  mutate(conf.low = estimate-2*std.error,
         conf.high = estimate+2*std.error) %>% 
  mutate(group = factor(group, 
                        levels = c("regular", "big"),
                        labels = c("Traditional Team Science", "BTS")))

@@

```{r fig-career, include = TRUE,  warning = FALSE, message = FALSE, fig.cap = "Average career length for big-team science authors. Larger dots indicate more variability in career length for authors by averaging the standard deviation in career length for each manuscript within a year. The data has been filtered to at least 10 publications in a year for this graph.", fig.width=10, fig.height=6}
ggplot(DF$career_time %>% 
         filter(SubjectArea != "Multidisciplinary") %>% 
         filter(num_papers >= 10) %>% 
         filter(Year < 2025) %>% 
         mutate(Team = factor(Team, 
                              levels = c("Regular", "Big"),
                              labels = c("Traditional Team Science", "BTS"))), 
       aes(Year, avg_career, color = Team)) +
  theme_classic(base_size = 15) + 
  geom_point(aes(size = sd_career)) +
  facet_wrap(~SubjectArea) + 
  theme(legend.position = "bottom") + 
  ylab("Average Career Length") +
  xlab("Year of Publication") + 
  scale_color_manual(values = c("#D55E00", "#0072B2"), name = "Team Size") + # Red and Blue from RdBu ends
  guides(size = "none")
  #scale_size_continuous(name = "Number Papers")

@@

Figure \@ref(fig:fig-career) portrays the average career length for
authors involved in traditional team science and BTS projects over time. Career
length was defined as the year of first publication minus the current
year, and higher numbers mean longer careers. The general pattern for
traditional team science and BTS projects is a decrease in average career length
over time. However, it appears that, in at least the last two decades,
BTS projects average a longer career length than traditional team science projects.
This trend is visually consistent across all four subject areas
examined.

@@

All slopes for both the average ($b_M$) and standard deviation
($b_{SD}$) of publication count were significantly different from zero,
indicating meaningful change over time in the types of researchers
contributing to publications across subject areas and team sizes (see
Figure \@ref(fig:fig-heatmap)). Most subject areas differed
significantly from one another within their team size, with the
exception of Health Sciences and Physical Sciences for traditional team science, whose $b_M$ values overlapped in their confidence intervals.
Across the remaining comparisons, Life Sciences showed the steepest
decline in average publication count over time for regular teams ($b_M$
= -0.0588), suggesting a shift toward including authors with fewer
publications. In contrast, the smallest change in publication count was
observed in Physical Sciences for big teams ($b_M$ = -0.0005),
indicating some stability scholar publication count when examining
diversity. Standard deviation slopes ($b_{SD}$) were generally low in
magnitude, with both positive and negative values depending on subject
area. This suggests some variation in the diversity of publication rates
across disciplines, with no uniform pattern of increasing or decreasing
diversity.

All subject areas showed significant differences between BTS and team
science teams in both $b_M$ and $b_{SD}$, as indicated by
non-overlapping confidence intervals. Traditional team science consistently
exhibited stronger negative slopes for average publication count than
BTS teams, reflecting a more pronounced trend toward authors with fewer
publications appearing over time. This result suggests that smaller teams are
increasingly composed of researchers with lower overall publication
counts, whereas BTS teams show a more muted shift. Effect sizes for
these models were smaller than those observed for career length, with
$R^2$ values ranging from .0335 to .0977. The strongest association was
observed in the Physical Sciences for big teams ($R^2$ = .0977), though
all models showed low-to-moderate predictive ability.

@@

Geographic patterns in authorship differed notably between BTS and team
science publications, as shown in the publication maps and mosaic charts
(Supplemental Figure \@ref(fig:fig-map-both)). BTS publications were overwhelmingly
concentrated in high-income countries, particularly the United States,
Western European nations (e.g., Germany, the United Kingdom, France, and
the Netherlands), and East Asian countries (e.g., China, Japan, and
South Korea). In contrast, traditional team science publications showed broader
geographic distribution, with relatively higher representation from
Latin America (e.g., Brazil, Mexico), South and Southeast Asia (e.g.,
India, Pakistan, Indonesia), and parts of Africa and the Middle East.
While both team types were led by traditionally defined Global North
institutions, the mosaic charts revealed that traditional team science 
included a more diverse range of countries contributing at moderate
levels. These patterns suggest that although BTS involves international
collaborations, it remains more centralized in historically dominant
research regions, whereas traditional team science may offer relatively greater
global inclusivity at a smaller scale.

To understand the change in representation diversity, we examined if the
number of regions in a publication is predicted by the year of
publication. Increasing diversity would be represented by a positive
slope, while decreasing diversity would be represented by a negative
slope. All slopes predicting geopolitical diversity over time were
significantly different from zero, indicating small but non-zero changes
in the number of regions represented on publications across disciplines
and team types. Additionally, all slopes differed significantly between
BTS and traditional team science publications, suggesting distinct patterns in the
evolution of international collaboration. Within BTS publications, Life
Sciences and Social Sciences showed statistically indistinguishable
trends in regional diversity over time, as did Social Sciences and
Health Sciences. In contrast, all other within-BTS comparisons differed
significantly. For traditional team science publications, all four disciplines
showed significantly different slopes, although the magnitudes of these
differences were relatively small. Overall, the results suggest modest
increases in geopolitical diversity in most disciplines, with a small
decline observed only in Physical Sciences within BTS publications ($b$
= -0.0129). Despite small effect sizes (all $R^2 < .012$), the
consistent differences between BTS and traditional team science point to
structural differences in how global participation is evolving across
large-scale versus more traditional collaborations.

@@

Across all decades and team sizes, North America and Northwestern Europe
consistently made up the majority of corresponding authors, as shown in
Figure \@ref(fig:fig-author-gpe). This pattern held even as total
team size increased, though the proportion of corresponding authors from
other regions (e.g., Asia, Latin America and the Caribbean, and Africa)
showed gradual increases over time. For traditional team science (2–10
authors), the dominance of North America and Western Europe in
corresponding author roles was particularly pronounced. In contrast, for
very large teams (100+ authors), regional diversity appeared somewhat
more balanced, with more visible contributions from Asia and other
regions among both corresponding and non-corresponding authors. However,
visual inspection suggests that corresponding author sets remained less
regionally diverse than the rest of the author team, particularly in
earlier decades. While representation from regions like Africa and Latin
America grew slightly among non-corresponding authors, they remained
minimally represented in lead authorship positions. Notably, Eastern
Asia’s representation increased more substantially over time, especially
in teams with 50 or more authors. These visual trends suggest persistent
regional disparities in leadership roles within scientific publications,
despite increasing global collaboration. Quantitative effect sizes
(Cramer’s *V*) and confidence intervals are reported in the following
section to determine the importance of these observed differences.

@@

This study expands on prior efforts to characterize Big Team Science
(BTS) by providing a systematic, field-wide analysis of authorship
composition across time, team size, and geography. While BTS efforts
have been increasingly promoted as vehicles for collaboration, scale,
and rigor [@adams2012; @uhlmann2019], questions remain about who gets
included, who leads, and how equitably credit is distributed. A key
contribution of our study is that it is the first to propose a
data-driven operationalization of "big" teams: defined here as
publications with 11 or more authors and contributions from at least 6
institutions, grounded in the empirical distribution of team sizes and
affiliations across millions of papers. By comparing BTS publications
with traditional team science publications across four major scientific domains, we
clarify how BTS compares with traditional team science and how it is
evolving. All areas of research show growth in the number of
publications and authors included on manuscripts, replicating previous
investigations [@sinatra2015; @wuchty2007; @hunter2008].

Next, we find that early-career researchers are increasingly represented
in both BTS and smaller teams. Across all disciplines, the average
career length of authors decreased significantly over time. Traditional team science exhibited even steeper declines in both average and variability of
career length than big teams, suggesting that smaller teams may be an
especially important entry point for early-career researchers. These
trends echo broader shifts in academia’s incentive structures, where
publishing early and often is increasingly required for career
advancement [@milojevic2014; @larivière2015]. Publication counts showed
similar but smaller effects. The average number of publications per
author declined over time in both big and small teams, with smaller
teams again showing more pronounced shifts. These findings support
claims that collaborative science is no longer dominated exclusively by
elite or high-output researchers [@milojevic2014], but may instead be
expanding to include contributors with more varied publication
histories.

Our findings build on previous research by also examining diversity in
author seniority and geopolitical affiliation. The growing participation
of early-career scholars over time suggests that big team science may be
increasingly accessible to a broader range of researchers, not just
senior or established scientists. This trend is interesting given the
challenges BTS projects can pose for non-permanent researchers: slow
publication timelines, uncertain publication outcomes, and fewer
incentives for non-corresponding authors. Yet, large teams may allow for
more distributed workloads and reduced individual time investment, which
could make them appealing even for early-career researchers. Moreover,
prior work has shown that publications from larger teams tend to receive
more citations and have broader impact [@larivière2015], which may
further incentivize early-career involvement despite the structural
risks. Globalization, the internet, and the focus on interdisciplinary
research are potentially driving forces behind our results, but,
hopefully, the results also point to a decline in scientific gate keeping
[@lu2007; @siler2015].

Our results confirm and extend prior observations that both traditional team science
and BTS are disproportionately concentrated in high-resource, highly
networked regions, namely, North America and Western Europe [@singh2023;
@adams2012; @sugimoto2017]. However, this study offers a more nuanced
picture. We observed modest increases in the geographic diversity of
authorship. Yet, lead authorship remained concentrated in a relatively
narrow set of regions. These findings parallel previous critiques of
global equity in scientific collaboration, where authors from the Global
South are often included in co-authorship lists but remain
underrepresented in leadership roles [@chan2011; @sumathipala2004].
Though Cramer’s *V* values reflecting geographic imbalance decreased
over time, especially for small and mid-sized teams, some asymmetries
persist in large teams, reinforcing concerns about exclusion even within
globally scoped research efforts [@abimbola2019].

@@

The limitations for this research are tied to the curation of the Scopus
dataset: the correct author affiliations, the correct author publication
information, and the correctly marked geopolitical entity. Scopus is a
carefully curated and large dataset, but these limitations must be kept
in mind when interpreting the results. Publication language diversity
was not investigated, and a previous study indicates that most
publications in big databases are in English [@albarillo2014].
Certainly, publications in non-English languages would improve the
statistics on diversity in scientific publishing - but the English
language barrier likely exists regardless of inclusion in databases
[@meneghini2007; @ramírez-castañeda2020]. Another limitation is the process of indexing across time has significantly changed with the availability of the internet. For example, in Figure \@ref(fig:fig-paper-time), traditional team science papers show a bump in the number of publications around approximately 1995, and the Web of Science came online in 1997, which included back indexing of journals not previous indexed.
